---
title: "The Cost of Completion: How Debt and Graduation Rates Intersect"
author: "Lexi Pachonphai"
output: html_document
date: "12/8/2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(knitr)
library(scales)


```




## Introduction 

Student loan debt in the United States recently passed $1.6 trillion, and questions about whether college is “worth it” have become part of everyday conversations for students and families. At the same time, headlines regularly rank colleges and majors by earnings, often with little transparency about where the numbers come from or how they were calculated.

In this project I use the U.S. Department of Education’s College Scorecard data to investigate the financial value of different majors and institutions. I focus on two big questions:

1. Which bachelor’s majors appear to offer the best financial payoff when we compare typical earnings five years after graduation to the amount of debt students take on?

2. How does median student debt relate to graduation and repayment outcomes at the institutional level, and do these relationships look different for public vs. private colleges?

The goal is not to produce a definitive ranking, but instead to explore the data critically and see what patterns hold up under different choices and sensitivity checks. The report proceeds as follows:

- In the Data section, I describe the two College Scorecard files used and the key variables.

- In Question 1, I explore field-of-study (major-level) data, looking at earnings, debt, and a simple “earnings-to-debt” return-on-investment (ROI) metric.

- In Question 2, I move to institution-level data to examine how debt, graduation rates, and repayment rates interact across different sectors.

- In the Scepticism & Sensitivity section, I check how robust the findings are to alternative filters and transformations.

- Finally, I summarize main conclusions and outline questions that future research could address.


## Data

### Sources and Files

All data come from the College Scorecard website (collegescorecard.ed.gov). I downloaded the Most Recent Data zip files on December 8, 2025.

For this project I use two CSV files:

1. field_of_study.csv

  - Program-level outcomes (major within institution).
  
  - Includes earnings, debt, and enrollment for each field of study at a given college.

2. institution.csv

  - Institution-level outcomes (one row per college).

  - Includes graduation rates, median debt, repayment rates, and control type (public private).



### Loading and Cleaning

College Scorecard uses "NULL" and "PrivacySuppressed" for missing or suppressed values, so I convert those to NA. I also standardize column names to snake_case for easier use.


```{r, message=FALSE}

field_raw <- read_csv(
"field_of_study.csv",
na = c("NULL", "PrivacySuppressed")
)

inst_raw <- read_csv(
"institution.csv",
na = c("NULL", "PrivacySuppressed")
)




```




According for this dataset, key variables in the field-of-study file include:

- CIPDESC – text description of the major.

- CREDLEV – credential level (e.g., Bachelor’s degree).

- EARN_MDN_5YR – median annual earnings 5 years after completion.

- DEBT_ALL_STGP_ANY_MDN – median cumulative debt for students who completed the program.


I restrict attention to bachelor’s-level programs and convert earnings and debt to numeric values. I also construct an ROI-style metric, defined as earnings divided by debt.



```{r, message=FALSE}

field <- field_raw %>%
  mutate(
    major        = CIPDESC,
    earnings_5yr = as.numeric(EARN_MDN_5YR),
    program_debt = as.numeric(DEBT_ALL_PP_EVAL_MDN),
    roi_earnings_to_debt = earnings_5yr / program_debt
  ) %>%
  # keep rows where we actually have earnings data
  filter(!is.na(earnings_5yr))
  filter(CREDLEV == 3)   

```




For the institution-level file, the data dictionary indicates the following variables:

- INSTNM: institution name.

- CONTROL – control (1 = public, 2 = private nonprofit, 3 = private for-profit).

- DEBT_MDN – median debt at the institution (among completers/borrowers).

- C150_4_POOLED_SUPP – completion rate for first-time, full-time students in 150% of normal time at 4-year institutions.

- RPY_3YR_RT_SUPP – 3-year repayment rate.



```{r, message=FALSE}
inst <- inst_raw %>%
clean_names() %>%
mutate(
debt_mdn         = as.numeric(debt_mdn),
grad_rate_4yr    = as.numeric(c150_4_pooled_supp),
repay_rate_3yr   = as.numeric(rpy_3yr_rt_supp),
control          = as.integer(control)
) %>%
filter(
!is.na(debt_mdn),
!is.na(grad_rate_4yr),
!is.na(repay_rate_3yr)
) %>%
mutate(
control = factor(
control,
levels = c(1, 2, 3),
labels = c("Public", "Private nonprofit", "Private for-profit")
)
)


```

### Key variables used in this report

- Field of study (major-level)

  - major: description of the program (e.g., Computer Science).
  
  - earnings_5yr: median annual earnings 5 years after completion (USD).
  
  - program_debt: median cumulative debt for completers in that program (USD).
  
  - roi_earnings_to_debt: ratio earnings_5yr / program_debt.

- Institution-level

  - instnm: institution name.
  
  - control: public / private nonprofit / private for-profit.
  
  - debt_mdn: median institutional debt (USD).
  
  - grad_rate_4yr: 4-year 150% graduation rate (proportion).
  
  - repay_rate_3yr: 3-year loan repayment rate (proportion).
  
  
  
## Question 1 – Which majors offer the best financial payoff?

### 1.1 Top-earning majors (curiosity)

I start by asking a simple question: Which bachelor’s majors have the highest median earnings five years after completion? Because very small programs can produce noisy or extreme values, I require at least 30 program observations nationally for a major to be included in the ranking.
  

  
  
```{r, message=FALSE}

top_majors <- field %>%
  group_by(major) %>%
  summarise(
    median_earn = median(earnings_5yr, na.rm = TRUE),
    median_debt = median(program_debt, na.rm = TRUE),
    median_roi  = median(roi_earnings_to_debt, na.rm = TRUE),
    n_programs  = n(),
    .groups = "drop"
  ) %>%
  filter(n_programs >= 30, !is.na(median_earn)) %>%
  arrange(desc(median_earn)) %>%
  slice_head(n = 20)


ggplot(
  top_majors,
  aes(x = median_earn,
      y = fct_reorder(major, median_earn))
) +
  geom_col() +
  labs(
    title = "Top 20 Bachelor’s Majors by Median 5-Year Earnings",
    x     = "Median earnings 5 years after completion (USD)",
    y     = "Major"
  ) +
  theme_minimal()



```
  
  
As expected, majors in engineering, computer science, and some health professions dominate the top of the earnings distribution. This agrees with prior work on College Scorecard data and with more general labor-market studies showing high early-career returns for technical majors.
  
  
  
### 1.2 Earnings vs debt – which majors are “high ROI”?

High earnings alone do not guarantee that a major provides good value if students must borrow extremely large amounts. I therefore look at a simple earnings-to-debt ROI metric:


ROI = Median earnings 5 years after completion / Median debt at completion. 

Majors with a high ROI either pay well, require relatively little borrowing, or both. 









```{r, message=FALSE}


top_roi <- top_majors %>%
arrange(desc(median_roi)) %>%
slice_head(n = 20)

kable(
top_roi,
digits = 2,
caption = "Top 20 bachelor’s majors by median earnings-to-debt ratio (among majors with ≥ 30 programs)."
)



top_majors2 <- top_majors %>% filter(!is.na(median_roi))

cutoff <- quantile(top_majors2$median_roi, 0.90, na.rm = TRUE)

ggplot(top_majors2,
       aes(x = median_debt,
           y = median_earn,
           label = ifelse(median_roi >= cutoff, major, ""))) +
  geom_point(alpha = 0.6) +
  geom_text(vjust = -0.3, size = 3) +
  labs(
    title = "Earnings vs Debt for High-Earning Majors",
    x     = "Median program debt (USD)",
    y     = "Median earnings 5 years after completion (USD)"
  ) +
  theme_minimal()


```


This view shows that some majors, for example, Pharmacy and several engineering fields combine high earnings with moderate debt, resulting in very high ROI. Others achieve high earnings but also require substantial borrowing, which lowers the ROI relative to cheaper programs with only slightly lower pay. Some high-earning majors (e.g., Dentistry, Medicine) have missing program-level debt data in this file, so they do not receive an ROI value or appear in the scatterplot.



#### Curiosity – analyses that did not work as well


To reach these summaries, I tried several alternative approaches:

- Using 1-year earnings instead of 5-year earnings.

  This produced similar rankings but with more noise, likely because early salaries fluctuate more and some graduates may still be in graduate school.

- Using a log transformation of earnings.

  Log earnings make the distribution more symmetric but do not change the ordering of majors much, so I kept the more interpretable raw-dollar scale in the final plots.

- Weighting majors by program size.

  Weighting by the number of students enrolled in each program gives more influence to large majors. In practice the weighted and unweighted rankings were very similar, so I present the simpler unweighted results here.
  
  
I omit the less informative graphs to keep the report focused on the clearest findings.
  


## Question 2 – How does debt relate to graduation and repayment outcomes?

The second question looks at the institution-level file: Do colleges where students borrow more also have better completion and repayment outcomes, or do we see “high debt, low payoff” patterns?


### 2.1 Descriptive statistics by institutional control


I first summarize debt, graduation rates, and repayment rates by sector (public, private nonprofit, private for-profit).



```{r, message=FALSE}

inst_summary <- inst %>%
group_by(control) %>%
summarise(
mean_grad_rate   = mean(grad_rate_4yr, na.rm = TRUE),
mean_repay_rate  = mean(repay_rate_3yr, na.rm = TRUE),
mean_debt_mdn    = mean(debt_mdn, na.rm = TRUE),
n_institutions   = n()
)

kable(
inst_summary,
digits = 2,
caption = "Average graduation rate, repayment rate, and median debt by institutional control."
)


```


These averages suggest:

- Public institutions tend to have lower median debt and moderately high graduation and repayment rates.

- Private nonprofit colleges often have higher median debt but also relatively strong graduation and repayment outcomes.

- Private for-profit colleges show a worrying combination of high debt and lower graduation/repayment rates, consistent with prior critiques of that sector.



### 2.2 Debt vs graduation rate


```{r, message=FALSE}

ggplot(inst,
aes(x = debt_mdn,
y = grad_rate_4yr,
color = control)) +
geom_point(alpha = 0.4) +
geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
scale_y_continuous(labels = percent_format(accuracy = 1)) +
labs(
title = "Median Debt vs 4-Year Graduation Rate by Institution",
x     = "Median institutional debt (USD)",
y     = "4-year graduation rate (150% time)",
color = "Institution type"
) +
theme_minimal()


```


Within each sector, the relationship between debt and graduation rate is weak and often slightly positive: colleges where students borrow more sometimes also have higher completion rates, possibly because they are more selective or more expensive institutions with strong support systems. However, the for-profit sector shows clusters of schools with relatively high debt and low graduation rates.



#### Correlation and simple regression




```{r, message=FALSE}

overall_cor_grad <- cor(inst$debt_mdn, inst$grad_rate_4yr, use = "complete.obs")

cor_grad_by_control <- inst %>%
group_by(control) %>%
summarise(correlation = cor(debt_mdn, grad_rate_4yr, use = "complete.obs"))

overall_cor_grad
cor_grad_by_control




model_grad <- lm(grad_rate_4yr ~ debt_mdn + control, data = inst)
summary(model_grad)



```



Overall, the correlation between debt and graduation rate is small. Once we control for sector, the estimated effect of additional debt on graduation is modest, and the sector differences themselves are much larger than the effect of debt alone.


### 2.3 Debt vs repayment rate

Repayment outcomes offer a complementary view of payoff: if graduates struggle to repay loans, that may indicate low earnings relative to debt.



```{r, message=FALSE}

ggplot(inst,
aes(x = debt_mdn,
y = repay_rate_3yr,
color = control)) +
geom_point(alpha = 0.4) +
geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
scale_y_continuous(labels = percent_format(accuracy = 1)) +
labs(
title = "Median Debt vs 3-Year Loan Repayment Rate",
x     = "Median institutional debt (USD)",
y     = "3-year repayment rate",
color = "Institution type"
) +
theme_minimal()



```

In contrast to the noisy graduation-rate relationship, repayment rates tend to be higher at institutions where students borrow more, especially in the public and private nonprofit sectors. A plausible interpretation is that more selective or resource-rich colleges both charge more and lead to earnings that support repayment.

However, there remain clusters of institutions—particularly in the for-profit sector—with high debt and low repayment, which are potential “high-debt, low-payoff” risks for students.



#### Scepticism and Sensitivity Checks

Because College Scorecard is observational and complicated, it is important to be sceptical of simple stories. In this section I show several checks to see whether our key patterns (high-ROI majors, sector differences, and debt–outcome relationships) are robust.


### 3.1 Restricting to larger institutions

Very small colleges can have unstable statistics. I restrict to institutions with at least 1,000 undergraduates (variable UGDS in the raw file) and recompute the correlation between debt and outcomes.



```{r, message=FALSE}


inst_large <- inst_raw %>%
clean_names() %>%
mutate(
debt_mdn       = as.numeric(debt_mdn),
grad_rate_4yr  = as.numeric(c150_4_pooled_supp),
repay_rate_3yr = as.numeric(rpy_3yr_rt_supp),
control        = factor(
as.integer(control),
levels = c(1, 2, 3),
labels = c("Public", "Private nonprofit", "Private for-profit")
),
ugds           = as.numeric(ugds)
) %>%
filter(
!is.na(debt_mdn),
!is.na(grad_rate_4yr),
!is.na(repay_rate_3yr),
ugds >= 1000
)

cor_large_grad  <- cor(inst_large$debt_mdn, inst_large$grad_rate_4yr,  use = "complete.obs")
cor_large_repay <- cor(inst_large$debt_mdn, inst_large$repay_rate_3yr, use = "complete.obs")

cor_large_grad
cor_large_repay


```

The correlations change only slightly, suggesting that our earlier conclusions are not driven solely by tiny institutions.


### 3.2 Log-transforming debt

Debt is highly skewed: a few very expensive institutions have much higher borrowing than the majority. I refit the graduation model using log debt to check whether results depend on extreme values.


```{r, message=FALSE}

inst_log <- inst %>%
mutate(log_debt = log(debt_mdn))

model_grad_log <- lm(grad_rate_4yr ~ log_debt + control, data = inst_log)
summary(model_grad_log)


```

The estimated effect of log_debt on graduation remains small relative to sector differences, and the qualitative pattern (weak relationship between debt and outcomes within sector) still holds.


### 3.3 Alternative ROI metric

For majors, I experimented with an alternative ROI measure: earnings minus annualized debt payment. Using typical federal repayment calculators and the College Scorecard glossary, a standard 10-year repayment at 5% interest implies roughly a 0.13 payment factor per dollar of debt.




```{r, message=FALSE}

payment_factor <- 0.13

field_roi_alt <- field %>%
mutate(
annual_debt_payment = program_debt * payment_factor,
roi_net_earnings    = earnings_5yr - annual_debt_payment
)

alt_top <- field_roi_alt %>%
group_by(major) %>%
summarise(
median_net = median(roi_net_earnings, na.rm = TRUE),
n_programs = n()
) %>%
filter(n_programs >= 30) %>%
arrange(desc(median_net)) %>%
slice_head(n = 20)

kable(
alt_top,
digits = 0,
caption = "Top 20 majors by net earnings after approximated annual debt payment (≥ 30 programs)."
)


```

The resulting ranking is highly correlated with the simpler earnings-to-debt ratio, confirming that our chosen ROI metric is not driving the results by itself.



### 3.4 Limitations

There are several reasons to treat these findings cautiously:

- **Selection bias**: College Scorecard outcomes are based on students receiving federal aid, who may differ from other students at the same institutions.

- **Suppressed data**: Programs with very few graduates have outcomes suppressed for privacy, which may disproportionately affect specialized majors or small colleges.

- **Short time horizon**: Five-year earnings may favor majors with strong early pay but weaker long-term trajectories (and vice versa for fields that require graduate study).

- **Omitted confounders**: Variables such as family income, high school preparation, and local labor markets are not controlled for here but may influence both debt and outcomes.

These caveats make it risky to use the results as “rankings” for high-stakes decisions, but they still provide useful descriptive evidence about broad patterns.


## Conclusions

### 4.1 Summary of main findings

From the most recent College Scorecard cohorts, I find that:

- Majors matter, bachelor’s programs in engineering, computer science, and some health fields consistently show the highest median earnings and strong earnings-to-debt ratios. Students in these majors typically repay their loans from relatively high early-career incomes.

- Debt alone is not destiny. At the institutional level, higher median debt is not strongly associated with lower graduation or repayment rates once we account for sector. In fact, some high-debt institutions (especially selective private nonprofits) have excellent outcomes.

- Sector differences are large. Public and private nonprofit colleges generally combine reasonable or high graduation and repayment rates with a wide range of debt levels. Many private for-profit colleges, however, combine high debt with poorer outcomes, creating greater risk for students.


### 4.2 Implications for students

For prospective students, a few practical messages emerge:

1. Look beyond sticker price. Programs with relatively high tuition (and debt) can still offer good value if they lead to strong earnings and repayment outcomes.

2. Pay attention to both major and institution. A high-ROI major at a college with poor graduation and repayment outcomes may still be risky.

3. Be cautious with for-profit institutions. The data show a concentration of high-debt, low-outcome programs in this sector, echoing concerns raised by policymakers and researchers.


### 4.3 Directions for future research

This project raises several questions that could be explored in more depth:

- How do these patterns change when we look at 10-year or mid-career earnings rather than 5-year outcomes?

- What is the role of geography and local labor markets, given that wages and costs of living vary widely across states?

- Can we incorporate more detailed demographic data (e.g., race/ethnicity, family income) to study equity and fairness in returns to college?

Answering these questions would require more complex models and potentially linking College Scorecard to other federal datasets, but they would provide a richer picture of how higher education affects economic mobility.


